Double trouble: Terrestrial ectortherm populations have both reduced opportunity for selection and physiological plasticity

### 2023-01-10 ms.rmd
UPdated MS with summary of results

---
title: "Reduced plasticity and opportunity for selection across terrestrial ectotherm populations"
bibliography: ./bib/refs.bib
csl: ./bib/ecology-letters.csl
output: 
  bookdown::word_document2:
    toc: no
    toc_depth: 6
    number_sections: false
    reference_docx: ./bib/template.docx
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
    bibliography: refs.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, tidy = TRUE, fig.width = 10)
## numbers >= 10^5 will be denoted in scientific notation,		  ## numbers >= 10^5 will be denoted in scientific notation,
  ## and rounded to 2 digits		  ## and rounded to 2 digits
  options(digits = 2)

@@

@@

Variable thermal environments are expected to result in strong selection pressures that result in adaptation or the evolution of phenotypic plasticity -- both of which are considered critical for population resilience to human-induced climate change [@Seebacher2015]. Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to debates over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015;@Gunderson2015; @Pinsky2019]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., see @Seebacher2015], it is unclear if the magnitude of physiological adjustment is sufficient. In addition, low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems

Theory suggests that phenotypic plasticity will evolve in variable, but predictable, habitats with empirical data supporting this prediction.

@@

### *Meta-Analysis*
We analysed our data using multilevel meta-analytic (MLMA) and meta-regression (MLMR) models in R (vers. `r paste0(R.Version()$major, ".", R.Version()$minor)`) using *brms* [vers. `r utils::packageVersion("brms")` @Burkner2018; @Burkner2017; @stan] and *metafor* [vers. `r utils::packageVersion("metafor")` @Viechtbauer2010].  We fit both Bayesian and frequentist approaches to ensure that our results were consistent, and to create orchard plots more easily [vers. 2.0, @Nakagawa2021c]. For our Bayesian models, we ran 4 MCMC chains, each with a warm-up of 1000 followed by 4000 sampling iterations keeping every 5 iterations for a total of 3200 samples from the posterior distribution. We used flat Gaussian priors for 'fixed' effects (i.e., $N(0, 10)$) and a student t-distribution for 'random' effects (i.e., $student_{t}(3, 0, 10)$). We checked that all MCMC chains were mixing and had converged (i.e., $R_{hat} = 1$).

#### *Multi-level Meta-analysis (MLMA) Models*
We first fit multi-level meta-analysis (MLMA) models (i.e., intercept only models) for both $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$, that included study, species, and phylogeny as random effects to account for non-independence. We also included trait as a random effect to account for trait variation within the data. Our MLMA models allowed us to partition the variation in $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ among these key sources while accounting for total sampling variance in each. This allowed us to calculate total heterogeneity [i.e., $I_{total}^2$; *sensu* @NakagawaSantos2012; @Noble2022] along with various $I^2$ metrics describing the proportion of variance explained by each random effect level [@NakagawaSantos2012].

@@

>>>>

---
title: "Double trouble: Terrestrial ectortherm populations have both reduced opportunity for selection and physiological plasticity"
bibliography: ./bib/refs.bib
csl: ./bib/ecology-letters.csl
output: 
  bookdown::word_document2:
    toc: no
    toc_depth: 6
    number_sections: false
    reference_docx: ./bib/template.docx
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
    bibliography: refs.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, cache = TRUE, warning = FALSE, tidy = TRUE, fig.width = 10)
## numbers >= 10^5 will be denoted in scientific notation,		  ## numbers >= 10^5 will be denoted in scientific notation,
  ## and rounded to 2 digits		  ## and rounded to 2 digits
  options(digits = 2)

@@

    # Taxa
      taxa_sum <- data_wide %>%  
              group_by(habitat, phylum, class) %>% 
              summarise(spp = length(unique(species_full)), n = n()) %>%  
              data.frame()

      taxa_sum_long <- data_long2 %>% group_by(habitat, phylum, class) %>% 
              summarise(spp = length(unique(species_full)), n = n()) %>%  
              data.frame()

@@

Variable thermal environments are expected to result in strong selection pressures that lead to adaptation or the evolution of phenotypic plasticity -- both of which are considered critical for population resilience to human-induced climate change [@Seebacher2015]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable. Despite this theoretical expectation, empirical support for this prediction is scant [], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity). Reversible forms of phenotypic plasticity, such as acclimatization, may not be expected to adhere to such predictions.

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to debates over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015;@Gunderson2015; @Pinsky2019]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., see @Seebacher2015], it is unclear if the magnitude of physiological adjustment is sufficient. In addition, low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems

@@

### *Meta-Analysis*
We analysed our data using multilevel meta-analytic (MLMA) and meta-regression (MLMR) models in R (vers. `r paste0(R.Version()$major, ".", R.Version()$minor)`) using *brms* [vers. `r utils::packageVersion("brms")` @Burkner2018; @Burkner2017; @stan] and *metafor* [vers. `r utils::packageVersion("metafor")` @Viechtbauer2010].  We fit both Bayesian and frequentist approaches to ensure that our results were consistent, and to create orchard plots more easily [vers. 2.0, @Nakagawa2021c]. In addition, Bayesian methods better protect against type I errors in the presence of complex sources of non-independence [@Nakagawa2021; @Song2021]. For our Bayesian models, we ran 4 MCMC chains, each with a warm-up of 1000 followed by 4000 sampling iterations keeping every 5 iterations for a total of 3200 samples from the posterior distribution. We used flat Gaussian priors for 'fixed' effects (i.e., $N(0, 10)$) and a student t-distribution for 'random' effects (i.e., $student_{t}(3, 0, 10)$). We checked that all MCMC chains were mixing and had converged (i.e., $R_{hat} = 1$).

#### *Multi-level Meta-analysis (MLMA) Models*
We first fit multi-level meta-analysis (MLMA) models (i.e., intercept only models) for both $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$, that included study, species, and phylogeny as random effects to account for non-independence. We also included trait as a random effect to account for trait variation within the data. Our MLMA models allowed us to partition the variation in $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ among these key sources while accounting for total sampling variance in each. This allowed us to calculate total heterogeneity [i.e., $I_{total}^2$; *sensu* @NakagawaSantos2012; @Noble2022] along with various $I^2$ metrics describing the proportion of variance explained by each random effect level [@NakagawaSantos2012]. We also present 95% prediction intervals which describe the expected distribution of effects from future studies [@Nakagawa2021c; @Noble2022].

@@

The final dataset included a total of `r f <- taxa_sum %>% filter(habitat == "f") %>% summarise(spp = sum(spp), k = sum(n)); f[1,1]` freshwater (fishes = `r taxa_sum %>% filter(habitat == "f" & class == "Actinopterygii") %>% summarise(sp = sum(spp))`, Molluscs = `r taxa_sum %>% filter(habitat == "f" & phylum == "Mollusca") %>% summarise(sp = sum(spp))`, Amphibians = `r taxa_sum %>% filter(habitat == "f" & class == "Amphibia") %>% summarise(sp = sum(spp))`, Reptiles = `r taxa_sum %>% filter(habitat == "f" & class == "Reptilia") %>% summarise(sp = sum(spp))`, Arthropods = `r taxa_sum %>% filter(habitat == "f" & phylum == "Arthropoda") %>% summarise(sp = sum(spp))`, and a single Crustacean and Nematode species), `r m <- taxa_sum %>% filter(habitat == "m") %>% summarise(spp = sum(spp), k = sum(n)); m[1,1]` marine (fishes = `r taxa_sum %>% filter(habitat == "m" & class %in% c("Actinopterygii", "Agnatha","Chondrichthyes")) %>% summarise(sp = sum(spp))`, Annelids = `r taxa_sum %>% filter(habitat == "m" & phylum == "Annelida") %>% summarise(sp = sum(spp))`, Molluscs = `r taxa_sum %>% filter(habitat == "m" & phylum == "Mollusca") %>% summarise(sp = sum(spp))`, Echinoderms = `r taxa_sum %>% filter(habitat == "m" & phylum == "Echinodermata") %>% summarise(sp = sum(spp))`, Reptiles = `r taxa_sum %>% filter(habitat == "m" & class == "Reptilia") %>% summarise(sp = sum(spp))`, Arthropods = `r taxa_sum %>% filter(habitat == "m" & phylum == "Arthropoda") %>% summarise(sp = sum(spp))`, and a single species of Crustacean, Cnidarian), and `r t <- taxa_sum %>% filter(habitat == "t") %>% summarise(spp = sum(spp), k = sum(n)); t[1,1]` terrestrial species (Annelids = `r taxa_sum %>% filter(habitat == "t" & phylum == "Annelida") %>% summarise(sp = sum(spp))`, Molluscs = `r taxa_sum %>% filter(habitat == "t" & phylum == "Mollusca") %>% summarise(sp = sum(spp))`, Arthropods = `r taxa_sum %>% filter(habitat == "t" & phylum == "Arthropoda") %>% summarise(sp = sum(spp))`, Reptiles = `r taxa_sum %>% filter(habitat == "t" & class == "Reptilia") %>% summarise(sp = sum(spp))` and Amphibians = `r taxa_sum %>% filter(habitat == "t" & class == "Amphibia") %>% summarise(sp = sum(spp))` along with a single Tardigrade species) (Fig. \@ref(fig:fig1)A). We had more data on acute thermal responses (n = `r trimmed <- data_long2 %>% filter(grepl("lnRR", name) & !is.na(effect_size) & !is.na(V) & !V == 0); sum1 = trimmed %>% group_by(name, type) %>% summarise(n = n()) %>% filter(type =="acute"); sum(sum1$n)`) compared to thermal responses after an acclimation period (n = `r sum2 <- trimmed %>% group_by(name, type) %>% summarise(n = n()) %>% filter(type =="acclim"); sum2$n`) (Fig. \@ref(fig:fig1)A).
